---
title: "student number: 25039479"
editor_options: 
  markdown: 
    wrap: 72
---

## Originality declaration

I, [**Ruijue Song**], confirm that the work presented in this assessment
is my own. Where information has been derived from other sources, I
confirm that this has been indicated in the work.

I used Gemini to help me with data analysis, and I will specifically
mark the places where I used it.

And I have extra time of Reasonable Academic Adjustments.

## Initial project scope

### Background

The gambling market in the UK is substantial, generating £14.2 billion
in 2020. The associated gambling-related harms are increasingly viewed
as a critical public health issue rather than merely a leisure activity.
In response to the growing scale of gambling harms, the UK Government
published the White Paper **"High stakes: gambling reform for the
digital age"** in 27 April 2023. This White Paper prioritized protecting
**vulnerable people** (e.g., children and young adults) from gambling
harms. Despite the legislative focus and the gambling-related harm
review in 2023, the spatial dimension of gambling harms is not taken
into consideration, such as whether the betting stores (bookmakers)
physically cluster in vulnerable communities.

Consequently, the bookmakers dataset extracted from OSM data provides a
critical window into the spatial distribution patterns. Acting as a
consultant of Greater London Authority (GLA), this report seeks to
understand what are the spatial patterns of gambling density in London,
thereby providing the missing spatial analysis to inform local licensing
and planning decisions.

### Research question and hypothesis

**Research question:** What are the spatial patterns of betting
stores(bookmakers) in London in 2025? Do they form statistically
significant high-density zones similar to one another (hotspots)?

**Hypothesis:**

-   $H_1$ (Alternative Hypothesis): Betting stores in London in 2025 are
    not randomly distributed but exhibit significant spatial clustering,
    with certain districts experiencing disproportionately high density
    compared to others. And there are hotspot zones in London.

-   $H_0$ (Null Hypothesis): Betting stores in London in 2025 are
    randomly distributed without showing any hotspot zones.

### Evaluation of research datasets

**dataset**: Bookmakers dataset extracted from OSM data

-   **Source &Nature:** Provide by OSM data, the dataset contains
    polygons of bookmakers locations in London, which has to be
    transformed into point data for following density calculation.

-   **CRS:** The raw data is in WGS 84(EPSG:4326). A key pre-processing
    step will involve transforming this to British National Grid
    (EPSG:27700) to enable valid metric spatial operations (e.g.,
    calculating density per $km^2$).

-   **Collection Method:** The dataset is extracted from **OpenStreetMap
    (OSM)**, which is a collaborative project where data is created and
    maintained by volunteers. Unlike official administrative registers
    (e.g., from the UK Gambling Commission), this dataset relies on
    **crowdsourced contributions**. As a crowdsourced dataset, it s
    subject to contributor bias and temporal lag. There is no guarantee
    that every betting store in London has been mapped. Newly opened
    stores could be missing and permanently closed shops may not have
    been deleted.

-   **Representation:** OSM coverage is often better in densely
    populated or affluent areas where more active contributors reside.
    Consequently, betting shops in deprived peripheral areas might be
    slightly underrepresented.

### Explanation of the data wrangling to undertake

1.  **CRS Transforming:** All spatial datasets (OSM bookmaker points,
    MSOA boundaries, London boundary) will be transformed to the British
    National Grid (EPSG: 27700), which is essential for accurate area
    calculations (required for density) and spatial distance
    measurements, avoiding distortions inherent in the raw WGS84
    geographic system.

2.  **Data Cleaning:** Remove the duplicates in bookmaker points dataset
    and select the related columns for following analysis.

3.  **Spatial Subsetting:** The OSM bookmaker points will be filtered to
    retain only those falling strictly within the study area (London
    MSOAs boundaries). This step is necessary to remove extraneous data
    points located in the corners of the bounding box, ensuring the
    analysis is strictly confined to the GLA jurisdiction.

4.  **Data Aggregation:** This is the most critical step. Betting shop
    points will be spatially joined to the MSOA polygons to calculate a
    raw count per spatial unit and these counts will be converted to
    density. Normalising by area is vital to mitigate the Scale Effect
    of the Modifiable Areal Unit Problem (MAUP). Without this,
    physically larger MSOAs in outer London might artificially seem to
    have more gambling issues simply due to their size.

### Justified plan for the spatial data analysis that will be required

1.  **Kernel Density Estimation (KDE):** Visualising Intensity

We firstly generate a KDE surface to from the cleaning betting shop
points to identify the intensity in London, which could provide us with
a initial understanding of bookmakers distribution and intensity.

2.  **Global Moran's I:** Testing for Global Clustering in London

Following data aggregation to MSOAs, the Global Moran’s I statistic will
be calculated for betting stores density.

Visual patterns in KDE can be vague and misleading. To rigorously answer
the research question, we must statistically test against the Null
Hypothesis of Complete Spatial Randomness (CSR). Global Moran's I is the
appropriate statistic for areal data (MSOAs) to quantify whether there
is clustering across the entire city, confirming if the observed
patterns are systematic rather than random noise.

3.  **Getis Ord** $G_i^*$: Identifying Local Hotspots

While Global Moran's I confirms if clustering exists in London, it
cannot tell us where. Getis Ord $G_i^*$ is specifically selected over
other local indicators (like Local Moran's I) because it explicitly
identifies "High-High" clusters (high density surrounded by high
density). This directly answers the second part of the research
question: "where are significant hotspots?". It also provides actionable
evidence for identifying specific neighbourhoods where gambling
availability has reached saturation levels, potentially requiring policy
intervention.

### Constraints & Limitations

1.  Data Reliability

The betting shop data is sourced from OpenStreetMap (OSM), which is
Volunteered Geographic Information. It may suffer from completeness
bias, where shops in less active contributor areas (often peripheral or
deprived) are under-mapped, potentially leading to an underestimation of
the density in the very areas of interest.

2.  The Modifiable Areal Unit Problem (MAUP)

The analysis relies on aggregating points to MSOAs. The results (e.g.,
identified hotspots) are dependent on these specific administrative
boundaries. Using different scales (e.g., boroughs) could yield
different clustering patterns or spatial correlations.

## Data Pre-processing

```{r}
library(tidyverse)
library(sf)
library(tmap)
library(ggplot2)
library(janitor)
library(osmdata)
library(spatstat) # Point Pattern Analysis (KDE)
library(spdep) # Spatial statistics (Global Moran's I and Local Getis Ord Gi)
```

### Data Loading

1.  Loading the MSOA boundary data Due to GitHub's file size limits
    (\>50MB), the raw MSOA boundary data is not included in this
    repository.

    URL of MSOA boudnary:
    <https://geoportal.statistics.gov.uk/datasets/3b3b1cb15a744855881b1489741b861b_0/about>

```{r}
MSOAs <- st_read("data/gambling_related_harms_data/Middle_layer_Super_Output_Areas_December_2021_Boundaries_EW_BFE_V8_-1517080999235121072/MSOA_2021_EW_BFE_V8.shp") %>% 
  st_transform(., 27700) #Ensure it is in British National Grid

st_crs(MSOAs) #Check its crs
```

2.  Loading London boundary data

URL:
<https://data.london.gov.uk/download/20od9/114d1137-e339-4b50-b409-124c17f4b59a/gla.zip>

```{r}
London <- st_read("data/gambling_related_harms_data/gla/London_GLA_Boundary.shp") %>% 
  st_transform(., 27700)

qtm(London)
```

Subsetting spatial units (MSOAs) to a study area (London)

```{r}
  #centroids of MSOAs across UK
  MSOAs_centroids <- st_centroid(MSOAs)

  #centroids within the London boundary. London is an sf object of the Greater London boundary 
  London_MSOAs <- MSOAs[st_within(MSOAs_centroids, London, sparse = FALSE), ]
  
  qtm(London_MSOAs)
```

3.  Extracting the bookmakers dataset from OSMs

```{r}
  Bounding_box <- getbb("London")
  #check it worked
  Bounding_box 
  
  #get the OSM data within the bounding box
  OSM_bookmaker <- Bounding_box %>%
  opq() %>%
  add_osm_feature(key = "shop", value = "bookmaker")%>%
  osmdata_sf()
  
  #extract the polygons of bookmakers
  bookmakers<-OSM_bookmaker$osm_polygons
  
  #convert to points
  bookmaker_points <- st_centroid(bookmakers)
```

```{r}
# CRS transforming for bookmaker points to British National Grid
bookmaker_points <- bookmaker_points %>% 
  st_transform(., 27700)

st_crs(bookmaker_points)
```

```{r}
summary(bookmaker_points)
```

Let us have a look at the points distribution firstly

```{r}
tmap_mode("plot")

# Set the base map
tm_shape(London_MSOAs)+
  tm_polygons(fill_alpha = 0.3)+
# Set the bookmakers dots
tm_shape(bookmaker_points)+
  tm_dots(fill = "blue", size = 0.2)
```

It is obvious that some points fall out of the London boundary, and we
have to remove them by subseeting in data wrangling part.

```{r}
#chek the column names and datatype list of bookmarker points data
Datatypelist <- bookmaker_points %>% 
  st_drop_geometry() %>% 
  summarise_all(class) %>% 
  pivot_longer(everything(),
               names_to = "All_variables",
               values_to = "Variable_class")
Datatypelist
```

Inspection of the dataset structure reveals that a significant number of
attribute columns contain NA values. As these attributes (e.g., address
details, opening hours) are not required for the proposed spatial
density analysis, the dataset will be streamlined.

In the data wrangling stage, we will select only the relevant columns
(e.g., name or osm_id) and the essential geometry column. This reduces
computational overhead and ensures a clean dataset for aggregation.

### Data Wrangling

#### 1. Data Cleaning: remove duplicates and select related columns

```{r}
# First we remove the duplicates in bookmaker points
bookmaker_points <- bookmaker_points %>% 
  clean_names() %>% 
  distinct()

# Then we keep only columns relevant to analysis: osm_id, name and geometry
bookmaker_selected <- bookmaker_points %>% 
  select(osm_id, name, geometry) %>% 
  # This step is to filter out rows with empty geometry (and I asked Gemini how to do it with sf)
  # Although it is rare in raw OSM data
  # I still take this step to ensure that subsequent spatial operations do not fail due to corrupt vector data
  filter(!st_is_empty(geometry)) 
```

#### 2. Spatial Subsetting

```{r}
# We remove any points falling outside London
bookmaker_sub <- bookmaker_selected[London_MSOAs,]
```

#### 3. Exploratory Mapping: bookmaker points in London

```{r}
# We plot to double check whether the points out of London removed
tmap_mode("plot")

tm_shape(London_MSOAs)+
  tm_polygons(fill_alpha = 0.3)+
tm_shape(bookmaker_sub)+
  tm_dots(fill = "blue", size = 0.2)
```

#### 4. Data Aggregation

Next, we calculates Density (count of betting stores per unit of area).
This reveals how concentrated the gambling is.

```{r}
points_sf_joined <- London_MSOAs %>% 
  clean_names() %>% 
  # First count points inside polygons
  # st_intersects returns a list of points found in each polygon
  # lengths() counts how many points in each unit
  mutate(n = lengths(st_intersects(., bookmaker_sub))) %>% 
  # Calculate the area of each spatial unit
  mutate(area = st_area(.)) %>% 
  # Calculate the density 
  # We multiply the density by 1,000,000 to convert it from per 1/m^2 to 1/km^2
  # Cause at first I used 1/m^2 to plot, find the density is so small (e.g., 0.00000008)
  # that is hard to read
  mutate(density = (n/area)*1000000) %>% 
  # Select related variables
  select(msoa21cd, msoa21nm, density, n)
```

We plot a choropleth map of betting stores density in London to perform
a preliminary visual inspection for potential 'saturation zones' or
clustering tendencies that warrant further investigation using global
and local spatial statistics.

```{r}
points_sf_joined <- points_sf_joined %>% 
  group_by(msoa21cd) %>% 
  summarise(density = first(density), msoa21nm = first(msoa21nm), bookmaker_count = first(n))

# Plot the choropleth map of bookmaker density
tm_shape(points_sf_joined) +
  tm_polygons(fill = "density",
              fill.scale = tm_scale_intervals(
                values = "brewer.blues",
                style = "jenks"),
              fill.legend = tm_legend(
                title = "Bookmaker Density in London",
                title.size = 0.85,
                size =0.8,
                poisson = tm_pos_out(
                  "right",
                  "center",
                  pos.v = "center"
                )
              ))
```

The map rejects a random distribution, revealing a distinct
core-periphery pattern with the highest densities ($> 5.9$ shops/$km^2$)
concentrated in Inner London. Significant 'islands' of high density
appear in suburban hubs (e.g., Croydon，Stratford)-I asked Gemini to
identify their names because I am not familiar with London districts.
These potential 'saturation zones' (approaching 9 shops/$km^2$) suggest
strategic clustering in high-footfall areas. In the context of the
Public Health England review, these 'saturation zones' represent areas
of hyper-accessibility, where the risk of gambling-related harm is
likely elevated due to the concentrated volume of betting opportunities.

While visually suggesting clustering, visual inspection is prone to
subjectivity. Points pattern and spatial autocorrelation analysis are
required to statistically confirm if these patterns are significant.

## Data Analysis

### Points Pattern Analysis

####Kernel Density Esitimation (KDE)

**Why KDE is appropriate while Ripley's K is not:** While Ripley's K is
valuable for quantifying the existence and scale of clustering (Global
Statistics), it do not identify the location of these clusters. KDE
shows us WHERE the highest density of betting stores is, which allows
for the precise geographic identification of 'risk environments' where
gambling availability is highest. Ripley's K only indicates the scale of
clustering instead of its specific location.

1.  Defining Observation Window (London boundary)

```{r}
# Set the London boundary as the window
window <- as.owin(London)
plot(window)
```

2.  Data Conversion (sf $\rightarrow$ ppp) The sf point object
    (bookmakers) cannot be used directly for density calculation. It
    must be converted into a Planar Point Pattern (ppp) object, which is
    the specific format required by the spatstat package.

```{r}
# Extract the coodinations of bookmakers
coords <- st_coordinates(bookmaker_sub)

# Converting sf to ppp
bookmaker_sub.ppp <- ppp(x = coords[,1],
                         y = coords[,2],
                          window = window)
```

3.  KDE

```{r}
# Run KDE
bookmaker_sub.ppp %>% 
  density(., sigma = 1000) %>% 
  plot()
```

This KDE surface reveals the continuous 'atmosphere' of gambling
availability. It identifies a 'hyper-saturated' core (Yellow) and
distinct polycentric 'flares' in Outer London (Red/Orange). These
peripheral hotspots (e.g., Croydon, Stratford) confirm that shops are
not randomly scattered but strategically anchored to high-footfall
commercial arteries.

### Spatial Autocorrelation

#### Global Moran's I

**Why Global Moran's I is appropriate:** We use Global Moran's I to
evaluate whether the spatial pattern of betting store density **across
London (Global)** is clustering. A **Queen’s Contiguity** spatial weight
matrix was used to conceptualise spatial relationships because it
captures all touching neighbours (sharing edges or corners). This is
particularly appropriate for modelling 'risk environments' in London's
irregular urban fabric, ensuring that the spatial spillover of gambling
availability across complex administrative intersections is fully
accounted for.

1.  Defining Wight Matrix: Queen Contiguity

```{r}
# Create Neighbor List (Queen Contiguity)
nb <- points_sf_joined %>% 
  poly2nb(., queen = T)

# Create Spatial Weights List
lw <- nb %>% 
  nb2listw(., style = "W", zero.policy = TRUE) # zero.policy=TRUE handles districts (islands) with no neighbors
```

2.  Global Moran's I

```{r}
# Run Global Moran's I test
moranI <- points_sf_joined %>% 
  pull(density) %>% 
  as.vector() %>% 
  moran.test(., lw)

moranI
```

The Global Moran’s I test yields a Moran’s I statistic of $0.165$ with a
Z-score (standard deviate) of $9.02$ and a p-value of $< 2.2e-16$,
revealing a positive spatial autocorrelation (clustering) and rejecting
the Null Hypothesis (CSR).

The exceptionally high Z-score (9.02) confirms that the observed
clustering is not random noise, but a structurally significant,
systematic phenomenon across London. It provides statistical evidence
supporting the 'predatory clustering' hypothesis, suggesting operators
systematically co-locate to create continuous 'risk environments' rather
than distributing randomly. To translate this city-wide finding into
actionable policy intelligence, a Local Getis-Ord $G_i^*$ analysis is
required to pinpoint exactly where these hotspots.

#### Local Getis Ord $G_i^*$

**Why Getis Ord** $G_i^*$ is appropriate: We choose Local Getis Ord
$G_i^*$ over Local Moran's I (LISA) as the study prioritizes identifying
'hotspot zones' rather than spatial outliers. Informed by the Public
Health England review, Gi\* specifically detects otspots (high density
surrounded by high density), which directly corresponds to contiguous
'risk environments' of hyper-accessibility.Coordinated policy
intervention is needed in these areas.

1.  Local Getis Ord Gi\*

```{r}
# Run Local Getis Ord Gi* test
Gi <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  localG(., lw)

head(Gi)
```

2.  Plot

```{r}
# Add Gi data to points_sf_joined for plot
points_sf_joined <- points_sf_joined %>%
  mutate(density_G = as.numeric(Gi))

# Plot the Gi density map
tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "density_G",
      fill.scale = tm_scale_intervals(values = "brewer.blues",
                                      midpoint = "NA",
                                      style = "jenks"),
       fill.legend = tm_legend(title="Gi*, Bookmaker Density in London",
                              title.size=0.85,
                              size=0.8,
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```

The map translates density into statistical certainty, with Gi Z-scores
peaking at 7.294, indicating these clusters are structurally entrenched
'hotspot zones' rather than random occurrences. The analysis isolates a
contiguous inner-London core and distinct polycentric satellite hotspots
(e.g., Croydon, Stratford). These "islands" of hyper-accessibility
amidst colder residential areas confirm strategic clustering in
high-footfall hubs.

Aligning with the Public Health England review, these High-High clusters
represent "frontline risk environments". This validates the "predatory
clustering" hypothesis, suggesting operators aggressively concentrate
supply to maximize cumulative impact. For the GLA, this provides the
specific evidentiary basis to designate "Cumulative Impact Zones" and
refuse new licenses in these saturated districts.

## Conclusion & Reflection

#### Policy Implications

Targeting 'Hotspot Zones, the spatial analysis successfully translates
the abstract concept of "gambling harm" into a concrete instrument for
policy intervention. Rather than applying a blanket "one-size-fits-all"
restriction, **policy should be spatially targeted**. In these
high-density hotspots, the policy should be stricter. New license
applications should be refused because these areas already have enough
betting stores. This would help stop the clustering that leads to higher
risks of gambling harm in these communities.

#### Reflection on Datasets and Methods

1.  **Data Issue:** The betting shop data comes from OpenStreetMap,
    which is created by volunteers. This might have sampling bias. For
    example, shops in poorer or outer areas might be missing because
    fewer volunteers map there.

2.  **Methodological Limits (MAUP):** The results depend on the
    boundaries we used (MSOAs). This is the Modifiable Areal Unit
    Problem (MAUP). If we used different zones (like Boroughs), the
    results might look different.
